LME-MPM applied to quasi-brittle fracture.

### 2020-05-24 2020_EFM_MPM_Eigensoftening.tex
20200524

% End glossaries
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
  The objective of this work is to introduce an alternative
  technique to address the fracture process of brittle and
  quasi-brittle materials under the \acrfull{mpm} 
  framework. With this purpose the eigensoftening algorithm, developed
  originally for the \acrfull{otm} approximation scheme, is extended
  to the \acrshort{mpm} with the aim of present a suitable alternative
  to the existing fracture algorithms developed for the
  \acrshort{mpm}. The good fitting in the predictions made by the
  eigensoftening algorithm against both analytical and experimental
  results proofs the well performance of the method under challenging
  loads.
\end{abstract}

\begin{keyword}
Quasi brittle fracture \sep Local-\textit{max-ent} approximation \sep
Material Point Method \sep Solid Dynamics
\end{keyword}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:1}
Presence of cracks are a violation of the continuity requirement in
\acrfull{fem}. To overcome it, numerous of numerical
artifacts has been proposed with the aim of reproducing such a
complex behaviour. These techniques vary from employing cohesive
approaches \cite{Barenblatt,Hilleborg_1976}, by adaptively inserting
cohesive elements \cite{Ortiz_1999,Pandolfi_2002,Ruiz_2000} at solid
elements boundaries, or handling arbitrary cracks paths by level set
representation of the fracture surface \cite{Belytschko_03}. The
simulation of fracture propagation in a more accurate and effective
way can be considered as one of the original drivers for developing
novel spatial discretization methods like meshfree techniques. Some
examples of it are the \acrfull{mpm}
\cite{Schreyer_2002,Chen_2002,Nairn_2003,Zhenmao_2005}, the
\acrfull{efgm} \cite{BELYTSCHKO_1995,BELYTSCHKO_2000,Zhuang_2012,Muthu_2013}, the
\acrfull{sph} \cite{Wang_2020,Wang_2019}, the \acrfull{otm} \cite{Li2010,Li_2012,Pandolfi_2013,Li_2015} or Peridynamics
\cite{HA_2011,RABCZUK_2017} among others. \\

Regarding \acrshort{mpm}, fracture can be described numerically in
two ways. One is the ``CRAcks with Material Points (CRAMP)''
\cite{Nairn_2003,Nairn_2006} and consist in to remove the restriction
of the single-valued velocity field close to the crack thorough two or
more sets of nodes. In this method, different labels are assigned to the material points
and nodes to distinguish if they are in the same side of the crack or
not. Under this approach crack surface is described with line segments
in 2D and triangle patches in 3D cases. The chosen criteria for crack
propagation is based on such parameters as energy release rate
analyzed by Tan \& Nairn (2002)\cite{Nairn_2002}, and the stress
intensity factor or the J-integral discussed by Guo \& Nairn
(2004)\cite{Nairn_2004}. The other approach is to introduce failed
material points to describe the crack evolution. In this method, the
formation of failed points describes the nucleation of cracks, and
thereafter its propagation and branching. Consequently, the position
of the crack does not need to be explicitly stated. These represents
significant advantages over the ``CRAMP''. Under this approach, the
prediction of failure evolution is computed with a decohesion model,
which has been discussed by Chen {\it et al.}\cite{Zhenmao_2005} and
Schreyer {\it et al.}\cite{Schreyer_2002}. And has been successfully
employed to simulate the fracture of brittle materials by Chen {\it et
  al.} \cite{Chen_2002,Chen_2003} and Sulsky \& Schreyer
\cite{Sulsky_2004}.\\

Similar to this approach, Schmidt {\it et al.}
\cite{Schmidt_2009} introduced the concept of eigenfracture, where
they approximate the crack set by means of eigen-deformations, wich
enable the material to develop displacement jumps at no cost of local
elastic energy. Later, the eigenerosion approach to brittle facture
was developed by Pandolfi {\it et al.}
\cite{Pandolfi_2012,Pandolfi_2013}. In this technique the erosion of
the material point means that each material point can be either intact or be completely failed or eroded and has no load bearing
capacity. This method has been successfully applied to simulate high
complex phenomena such dynamic fragmentation of metals
\cite{Li_2015}. Recently, Zhang {\it et al.}
\cite{Zhang_EE_2020} adopted the eigenerosion to resolve the dynamic
fracture of brittle materials in the \acrshort{mpm}
framework. Nonetheless, in quasi-brittle materials simulations
performed with this approach exhibit an overestimation of tensile
stress and the strain peaks. And conventional \acrshort{mpm}/\acrshort{gimp} always
faces stress oscillations.\\

To overcome the limitations observed in the EigenMPM
\cite{Zhang_EE_2020}. The present research propose the eigensoftening
algorithm developed by Navas {\it et al.}
\cite{Navas_2017_ES,Navas2018a} for the \acrshort{otm} framework and
engineered for quasi-brittle materials. Inspired in the concept of the
crack band model \cite{Bazant83}, since energy dissipation is thorough
the softened (or failed) volume, it is able to capture the gradual rather than abrupt
dissipation of the fracture energy. Second, to mitigate stress
oscillations the \acrfull{lme} approximation technique
\cite{Arroyo2006} to deal with tensile instabilities, such those that
occurs when particles crosses an element boundary.\\

@@

The aim of this section is to describe and introduce some special
techniques required to face the fracture problem under the \acrshort{mpm}
framework. In consequence it is structured as follows: first in
\ref{sec:2.1} the \acrfull{npc} algorithm for the \acrshort{mpm}
will be exposed, next the \acrshort{lme} approximants are
introduced in \ref{sec:2.2} as an accurate alternative technique to
interpolate data between particles and nodes, and finally fracture
algorithms based in the eigendeformation concept are presented in \ref{sec:2.3}.

\subsection{The \acrshort{mpm} time integration : A \acrlong{npc}  scheme}
\label{sec:2.1}
The \acrshort{mpm}~\cite{Sulsky1994} is a meshfree Lagrangian-Eulerian
method where moving material points, henceforth particles, carries on
all the physical information of the local state ($\sigma_p,
\varepsilon_p$) and a set of fixed background nodes are introduced to compute the balance of momentum equation. Since the \acrshort{mpm} 
possesses the advantages of both Lagrangian and Eulerian descriptions,
no element distortion takes place in the \acrshort{mpm}. Therefore it
is an appropriate and efficient method to solve problems with moving
discontinuities such as fracture evolution.\\
Without loosing generality, the \acrshort{mpm} algorithm
can be described with three main steps: (i) a variational recovery
process, where particle data is projected to the grid nodes, (ii) an
Eulerian step, where balance of momentum equation is expressed as a
nodal equilibrium equation in a FEM-like procedure, and finally
(iii) a Lagrangian advection of the particles. In the present research
a explicit predictor-corrector time integration scheme is adopted. The
purposes of this choice is motivated due its proved robustness and
stability for dynamic computations. In the first stage, the nodal
velocity predictor is computed following \eqref{eq:Predictor-velocity}, 
\begin{equation}
  \label{eq:Predictor-velocity}
  \vec{\tilde{v}}_I^{k+1} = \frac{ N_{Ip}^{k} m_p (\vec{v}_p^k + (1 - \gamma)\ \Delta t\ \vec{a}_p^k)}{\tens{m}_I^{k+1}}\ .
\end{equation}
This way of computing the nodal predictor is both numerically stable
and minimise the computational effort. Once nodal velocities are
obtained, the essential boundary conditions are imposed. After that, a
Eulerian phase is computed in the set of nodes in a
\acrshort{fem}-like way, where nodal forces $\vec{f}_{I}^{k+1}$ are
computed thorough the equilibrium equation. Next the nodal velocities
are corrected in a \textit{corrector} stage,
\begin{equation}
  \label{eq:Corrector-velocity}
  \vec{v}_{I}^{k+1} = \vec{v}_{I}^{pred} + \gamma\ \Delta t\
  \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}\ .
\end{equation}
Finally updated the particles are advected in the Lagrangian stage as
\begin{align}
  \label{eq:Update-lagrangian-pce}
        &\vec{a}_p^{k+1} =
          \frac{N_{Ip}^k\vec{f}_{I}^{k}}{\tens{m}_I^k}\ ,\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + \Delta t\
        \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\ ,\\
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
         N_{Ip}^k\ \vec{v}_{I}^{k} +
        \frac{1}{2}\Delta t^2\ \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\ .
\end{align}
The complete pseudo-algorithm it is summarised in \ref{sec:expl-pred-corr}.

@@

is thorough a Newton-Raphson procedure, or in more challenging

@@

As noticed in \cite{Zhang_EE_2020}, the EigenMPM have two important
shortcomings. The first is the presence of stress instabilities even
with \acrshort{gimp} shape functions. To overcome it, \acrshort{lme}
approximants are introduced as an
alternative to the existing interpolation techniques. Its
impressive performance mitigating spurious stress oscillations
\cite{Wobbes2020} under the \acrshort{mpm} framework help to enhance
notoriously the quality of the results as we can see in Section
\ref{sec:3.1}. Where both interpolation techniques
are compared by carry out an eigenerosion simulation. The second
limitation is concerning its inability to simulate properly quasi-brittle
fracture \cite{Navas_2018_ES}. It can be solved thorough the
eigensoftening algorithm described in Section \ref{sec:2.3}. A proof
of it is exposed in Section \ref{sec:3.3}, where experimental results
of a drop-weight impact test are compared with those produced by
eigensoftening computations.

@@

\section{Conclusions}
\label{sec:4}
Based on the EigenMPM approach to fracture proposed by Zhang {\it et al.}
\cite{Zhang_EE_2020}, we have proposed two addition improvements which
solves the shortcomings presented in the original concept. First by
bring to the \acrshort{mpm} framework the concept of eigensoftening,
which allow to consider gradual failure process of each material
point. This concept can be compared with a cohesive law in the context
of cohesive fracture. Furthermore, more elaborated damage curves can
be designed to describe the failure process in more sophisticated
materials. In this direction, an interesting future improvement of the
algorithm would be the ability to simulate of crack patterns in
composite material such carbon fiber. This leads a challenging
research topic. Secondly, \acrshort{lme} is employed to remove from
the solution those wiggles due to inaccuracies in the stress
integration. This interpolation technique could help to improve
results in fracture propagation by adapting it thorough the
deformation gradient. \textit{A priori} this will enhance the already
impressive localisation properties of eigensoftening and eigenerosion
algorithms.

>>>>

% End glossaries
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\etal}{
  \textit{et al.}
}

@@

\begin{abstract}
  The objective of this work is to introduce an alternative
  technique to address the fracture process of brittle and
  quasi-brittle materials under the \acrfull{mpm} 
  framework. With this purpose the eigensoftening algorithm, developed
  originally for the \acrfull{otm} approximation scheme, is extended
  to the \acrshort{mpm} with the aim of presenting a suitable alternative
  to the existing fracture algorithms developed for the
  \acrshort{mpm}. The good fitting in the predictions made by the
  eigensoftening algorithm against both analytical and experimental
  results proofs the well performance of the method when challenging
  applications are to modeled.
\end{abstract}

\begin{keyword}
Quasi brittle fracture \sep Local-\textit{Max-Ent} approximation \sep
Material Point Method \sep Solid Dynamics
\end{keyword}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:1}
The presence of cracks is a violation of the continuity requirement of the
\acrfull{fem}. To overcome this issue, numerous numerical
artifacts have been proposed with the aim of reproducing such a
complex behaviour. These techniques vary from the employment of cohesive
approaches~\cite{Barenblatt,Hilleborg_1976}, \PNA{to the gradual insertion of}
cohesive elements \cite{Ortiz_1999,Pandolfi_2002,Ruiz_2000} at solid
elements boundaries, or handling arbitrary crack paths by level set
representation of the fracture surface~\cite{Belytschko_03}. The
simulation of fracture propagation in a more accurate and effective
way can be considered as one of the \PNA{ongoing goals} of the development of novel spatial discretization methods like meshfree techniques. Some
examples of it are the \acrfull{mpm}
\cite{Schreyer_2002,Chen_2002,Nairn_2003,Zhenmao_2005}, the
\acrfull{efgm} \cite{BELYTSCHKO_1995,BELYTSCHKO_2000,Zhuang_2012,Muthu_2013}, the
\acrfull{sph} \cite{Wang_2020,Wang_2019}, the \acrfull{otm} \cite{Li2010,Li_2012,Pandolfi_2013,Li_2015} or Peridynamics
\cite{HA_2011,RABCZUK_2017} among others. \\

Regarding \acrshort{mpm}, fracture \PNA{has been treated numerically through two different ways}. On the one hand, the ``CRAcks with Material Points (CRAMP)'' is proposed~\cite{Nairn_2003,Nairn_2006}. \PNA{The basis of this methodology consist of the removal of} the restriction
of the single-valued velocity field close to the crack through, at least, two sets of nodes. In this method, different labels are assigned to the material points
and nodes to distinguish if they are in the same side of the crack or
not. Under this approach, crack surface is described with line segments
in 2D and triangle patches in 3D cases. The chosen criteria for crack
propagation is based on such parameters as energy release rate
analyzed by Tan \& Nairn (2002)\cite{Nairn_2002}, and the stress
intensity factor or the J-integral discussed by Guo \& Nairn
(2004)\cite{Nairn_2004}. 
On the other hand, the \PNA{foundation} of the second approach is the introduction of failed
material points to describe the crack evolution. In these methods, the
formation of failed points describes the nucleation of cracks, and
thereafter its propagation and branching. Consequently, the position
of the crack does not need to be explicitly stated. This fact represents
significant advantages over the ``CRAMP''. Under this approach, the
prediction of failure evolution is computed with a decohesion model,
which has been discussed by Chen {\it et al.}\cite{Zhenmao_2005} and
Schreyer {\it et al.}\cite{Schreyer_2002}. Successful simulations are provided within the framework of the fracture of brittle materials by Chen {\it et
  al.} \cite{Chen_2002,Chen_2003} and Sulsky \& Schreyer
\cite{Sulsky_2004}.\\

Similar to this approach, Schmidt {\it et al.}
\cite{Schmidt_2009} introduced the concept of eigenfracture, where
crack sets are approximated by means of eigen-deformations, which
enable the material to develop displacement jumps at no cost of local
elastic energy. Later, the eigenerosion approach to brittle facture
was carried out by Pandolfi {\it et al.}
\cite{Pandolfi_2012,Pandolfi_2013}. \PNA{In this technique, the concept of ``erosion'' of
the material point is depicted, being the material point able to remain intact or become completely failed or eroded, when no loading
capacity is bore}. This method has been successfully applied to simulate high
complex phenomena such as dynamic fragmentation of metals
\cite{Li_2015}. Recently, Zhang {\it et al.}
\cite{Zhang_EE_2020} adopted the eigenerosion to resolve the dynamic
fracture of brittle materials in the \acrshort{mpm}
framework. Nonetheless, when quasi-brittle materials are simulated
though this approach, the results exhibit an overestimation of tensile
stress and strain peaks. Furthermore, since the work of Zhang {\it et al.}
\cite{Zhang_EE_2020} is made with conventional \acrshort{mpm}/\acrshort{gimp}, stress oscillations are to be faced.\\

To overcome the limitations observed in the EigenMPM
\cite{Zhang_EE_2020}, the present research proposes the eigensoftening
algorithm developed by Navas {\it et al.}
\cite{Navas_2017_ES,Navas2018a} for the \acrshort{otm} framework and
engineered for quasi-brittle materials. Inspired in the concept of the
crack band model \cite{Bazant83}, since energy dissipation is performed through the softened (or failed) volume, this methodology is able to capture the gradual rather than abrupt
dissipation of the fracture energy. \PNA{Moreover, in order to mitigate stress
oscillations as well as to deal with tensile instabilities, such those that
occurs when particles crosses an element boundary, the work of Molinos~\etal~\cite{Molinos2020} is followed. The traditional \acrshort{mpm} technique is enhanced through the \acrfull{lme} approximation technique
\cite{Arroyo2006} for the spatial discretization as well as a explicitly \acrshort{mpm} \acrfull{npc} scheme is derived for the time discretization.}\\

@@

The aim of this section is to describe and introduce some special
techniques required to face the fracture problem under the \acrshort{mpm}
framework. In consequence, this section is structured as follows: first, in
\ref{sec:2.1}, the \acrfull{npc} algorithm for the \acrshort{mpm}
will be exposed; next the \acrshort{lme} approximants are
introduced in \ref{sec:2.2} as an accurate alternative technique to
interpolate data between particles and nodes, and finally fracture
algorithms based on the eigendeformation concept are presented in \ref{sec:2.3}.

\subsection{The \acrshort{mpm} time integration : A \acrlong{npc}  scheme}
\label{sec:2.1}
The \acrshort{mpm}~\cite{Sulsky1994} is a meshfree Lagrangian-Eulerian
method where moving material points, henceforth in this research, particles will carry on
all the physical information of the local state ($\sigma_p,
\varepsilon_p$) and a set of fixed background nodes will be introduced to compute the balance of momentum equation. Since the \acrshort{mpm} 
possesses the advantages of both Lagrangian and Eulerian descriptions,
no element distortion takes place in the \acrshort{mpm}. Therefore, it
is an appropriate and efficient method to solve problems with moving
discontinuities such as fracture evolution.\\
Without loosing generality, the \acrshort{mpm} algorithm
can be described with three main steps: (i) a variational recovery
process, where particle data is projected to the grid nodes, (ii) an
Eulerian step, where balance of momentum equation is expressed as a
nodal equilibrium equation in a FEM-like procedure, and finally
(iii) a Lagrangian advection of the particles. In the present research,
an explicit predictor-corrector time integration scheme is adopted. The
purposes of this choice is motivated due to its proved robustness and
stability for dynamic computations. In the first stage, the nodal
velocity predictor is computed following \eqref{eq:Predictor-velocity}: 
\begin{equation}
  \label{eq:Predictor-velocity}
  \vec{\tilde{v}}_I^{k+1} = \frac{ N_{Ip}^{k} m_p (\vec{v}_p^k + (1 - \gamma)\ \Delta t\ \vec{a}_p^k)}{\tens{m}_I^{k+1}}\ .
\end{equation}
\PNA{EXPLICAR NOTACION, CUIDADO CON EL GAMMA} This way of computing the nodal predictor is numerically stable
as well as minimises the computational effort. Once nodal velocities are
obtained, the essential boundary conditions are imposed. After that, an
Eulerian phase is computed in the set of nodes in a
\acrshort{fem}-like way, where nodal forces $\vec{f}_{I}^{k+1}$ are
computed through the equilibrium equation. Next the nodal velocities
are corrected in a \textit{corrector} stage:
\begin{equation}
  \label{eq:Corrector-velocity}
  \vec{v}_{I}^{k+1} = \vec{v}_{I}^{pred} + \gamma\ \Delta t\
  \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}\ .
\end{equation}
Finally, in order to update the particles, those are advected in the Lagrangian stage as
\begin{align}
  \label{eq:Update-lagrangian-pce}
        &\vec{a}_p^{k+1} =
          \frac{N_{Ip}^k\vec{f}_{I}^{k}}{\tens{m}_I^k}\ ,\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + \Delta t\
        \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\ ,\\
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
         N_{Ip}^k\ \vec{v}_{I}^{k} +
        \frac{1}{2}\Delta t^2\ \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\ .
\end{align}
The complete pseudo-algorithm it is summarised in \ref{sec:expl-pred-corr}.

@@

is through a Newton-Raphson procedure, or in more challenging

@@

As noticed in \cite{Zhang_EE_2020}, the EigenMPM have two important
shortcomings. The first is the presence of stress instabilities even
with \acrshort{gimp} shape functions. To overcome it, \acrshort{lme}
approximants are introduced as an
alternative to the existing interpolation techniques. Its
impressive performance mitigating spurious stress oscillations
\cite{Wobbes2020} under the \acrshort{mpm} framework help to enhance
notoriously the quality of the results as we can see in Section
\ref{sec:3.1}. Where both interpolation techniques
are compared by carry out an eigenerosion simulation. The second
limitation is concerning its inability to simulate properly quasi-brittle
fracture \cite{Navas_2018_ES}. It can be solved through the
eigensoftening algorithm described in Section \ref{sec:2.3}. A proof
of it is exposed in Section \ref{sec:3.3}, where experimental results
of a drop-weight impact test are compared with those produced by
eigensoftening computations.

@@

\section{Conclusions}
\label{sec:4}
Based on the EigenMPM approach to fracture proposed by Zhang {\it et al.}
\cite{Zhang_EE_2020}, we have proposed two addition improvements which
solves the shortcomings presented in the original concept. First by
bring to the \acrshort{mpm} framework the concept of eigensoftening,
which allow to consider gradual failure process of each material
point. This concept can be compared with a cohesive law in the context
of cohesive fracture. Furthermore, more elaborated damage curves can
be designed to describe the failure process in more sophisticated
materials. In this direction, an interesting future improvement of the
algorithm would be the ability to simulate of crack patterns in
composite material such carbon fiber. This leads a challenging
research topic. Secondly, \acrshort{lme} is employed to remove from
the solution those wiggles due to inaccuracies in the stress
integration. This interpolation technique could help to improve
results in fracture propagation by adapting it through the
deformation gradient. \textit{A priori} this will enhance the already
impressive localisation properties of eigensoftening and eigenerosion
algorithms.
